---
title: "Multivariate Analysis of Spatial Inequality in England and Wales"
author: "Lance Pangilinan"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

# Introduction

*(Narrative as per accompanying PDF report.)*

# Data Overview

## Setup (Working Directory, Libraries, etc.)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Working Directory

```{r setwd}
# Clear workspace and set working directory
rm(list=ls())
setwd("~/Downloads/Multivariate_Stats")
```

### Libraries

```{r libraries}
# Load required packages for analysis
library(aplpack)    # Hierarchical clustering tools
library(fpc)        # Cluster analysis functions
library(cluster)    # Clustering algorithms
library(ape)        # Dendrogram plotting
library(amap)       # Distance metrics

library(MASS)       # MANOVA and multivariate analysis
library(biotools)   # Multivariate visualization

library(psych)      # Factor analysis tools
library(car)        # Companion to applied regression
library(corrplot)   # Correlation plots
library(EFAtools)   # Exploratory factor analysis

library(heplots)    # HE plots for multivariate data
```

### Load and Prepare Data

```{r data}
# Read dataset and select relevant variables
data <- read.csv("Housing_UK_2015.csv")
variables <- c("town", "all_price_2015", "housepricechange", "households_owned", 
               "households_rent_private", "crime_rank", "prop_health_vuln", 
               "noqual", "student_prop_16_74", "net_commuting", "prop_manu")

house <- data[, variables]
rownames(house) <- house[, "town"]
house <- house[,-1]

# Standardize variables
house2 <- scale(house)
```

# Descriptive Statistics

```{r sumstat}
# Summary statistics for each variable
summary(house)
```

```{r cqplot}
# Chi-square quantile plot for multivariate normality check
cqplot(house, main = "Chi-Square Quantiles for 2016 UK City Housing Data")
```

# Factor Analysis

```{r corrplot}
# Display raw correlations and plot correlation matrix
round(cor(house2),2)
corrplot.mixed(cor(house2), lower.col = "black", upper = "ellipse", tl.col = "black", 
               number.cex = .7, tl.pos = "lt", tl.cex = .7, 
               p.mat = cor.mtest(house2, conf.level = .95)$p, sig.level = .05)
```

```{r kmo}
# Check sampling adequacy for factor analysis
KMO(as.matrix(house2))
```

```{r pca}
# Perform Principal Component Analysis (PCA)
comp1 <- prcomp(house2, scale. = TRUE)

# Display PCA summary (variance explained and eigenvalues)
summary.PCA.JDRS <- function(x){
  sum_JDRS <- summary(x)$importance
  sum_JDRS[1, ] <- sum_JDRS[1, ]^2
  attr(sum_JDRS, "dimnames")[[1]][1] <- "Eigenvals (Variance)"
  sum_JDRS
}
round(summary.PCA.JDRS(comp1), 2)

# Display PCA loadings
round(comp1$rotation, 2)

# Scree plot of PCA
screeplot(comp1, type = "lines", col = "red", lwd = 2, pch = 19, cex = 1.2, 
          main = "Scree Plot of Standardized 2015 UK Housing Data")
```

*Elbow suggests retaining 2 principal components (latent factors).*

## Maximum Likelihood Estimation (MLE)

```{r mle}
# Conduct factor analysis using maximum likelihood estimation
factor <- factanal(house2, factors = 2, rotation = "varimax")
factor
```

```{r comparison}
# Assess fit: reproduced vs. observed correlation matrix
repro2 <- factor$loadings %*% t(factor$loadings)
resid2 <- cor(house2) - repro2
round(resid2, 2)

# Calculate Root Mean Square Residual (RMSR)
len <- length(resid2[upper.tri(resid2)])
(RMSR2 <- sqrt(sum(resid2[upper.tri(resid2)]^2)/len))

# Proportion of residuals > 0.05
paste0(round(sum(abs(resid2[upper.tri(resid2)]) > 0.05)/len*100), "%")
```

## Principal Axis Factoring (PAF)

```{r PAF}
# Conduct factor analysis using principal axis factoring
factor2 <- fa(house2, nfactors = 2, rotate = "varimax", fm = "pa")
factor2
```

```{r}
# Assess fit for PAF
repro2 <- factor2$loadings %*% t(factor2$loadings)
resid2 <- cor(house2) - repro2
round(resid2, 2)

# Calculate RMSR
len <- length(resid2[upper.tri(resid2)])
(RMSR2 <- sqrt(sum(resid2[upper.tri(resid2)]^2)/len))
paste0(round(sum(abs(resid2[upper.tri(resid2)]) > 0.05)/len*100), "%")
```

## Iterative PCA

```{r}
# Conduct iterative PCA with quartimax rotation
factor3 <- fa(house2, nfactors = 2, rotate = "quartimax", SMC = FALSE, fm = "pa")
factor3
```

```{r}
# Assess fit for iterative PCA
repro3 <- factor3$loadings %*% t(factor3$loadings)
resid3 <- cor(house2) - repro3
round(resid3, 2)

# RMSR and proportion of residuals > 0.05
len <- length(resid3[upper.tri(resid3)])
RMSR3 <- sqrt(sum(resid3[upper.tri(resid3)]^2)/len)
RMSR3
paste0(round(sum(abs(resid3[upper.tri(resid3)]) > 0.05)/len*100), "%")
```

### Loading Plots

```{r loading}
# Visualize factor loadings (MLE)
plot(factor$loadings, pch = 18, col = 'red')
abline(h = 0)
abline(v = 0)
text(factor$loadings, labels = names(house), cex = 0.8)
```

```{r loading}
# Visualize factor loadings (Iterative PCA)
plot(factor3$loadings, pch = 18, col = 'red')
abline(h = 0)
abline(v = 0)
text(factor3$loadings, labels = names(house), cex = 0.8)
```

# Hierarchical Clustering

## Clustering 1 – Euclidean Distance & Complete Linkage

```{r clustering1-complete, echo=TRUE, message=FALSE}
# Compute Euclidean distance matrix
dist_euc <- dist(house2, method = "euclidean")

# Perform hierarchical clustering using complete linkage
hc_complete <- hclust(dist_euc, method = "complete")

# Plot dendrogram with 6 clusters highlighted
plot(hc_complete,
     labels = rownames(house2),
     cex = 0.42,
     xlab = "",
     ylab = "Distance",
     main = "Euclidean Distance + Complete Linkage")
rect.hclust(hc_complete, k = 6)
```

```{r clustering1-evaluate, echo=TRUE}
# Load evaluation function and evaluate clustering
source("https://raw.githubusercontent.com/jreuning/sds363_code/refs/heads/main/HClusEval3.R.txt")
hclus_eval(house2,
           dist_m = 'euclidean',
           clus_m = 'complete',
           plot_op = TRUE,
           print_num = 15)
```

```{r clustering1-boxes, echo=TRUE}
# Re-plot dendrogram with cluster boundaries for reporting
plot(hc_complete,
     labels = rownames(house2),
     cex = 0.42,
     xlab = "",
     ylab = "Distance",
     main = "Clustering of Towns")
rect.hclust(hc_complete, k = 6)
```

## Clustering 2 – Euclidean Distance & Ward’s Method

```{r clustering2-ward, echo=TRUE, message=FALSE}
# Compute Euclidean distance matrix
dist_euc2 <- dist(house2, method = "euclidean")

# Perform hierarchical clustering using Ward's method
hc_ward <- hclust(dist_euc2, method = "ward.D")

# Plot dendrogram with 6 clusters highlighted
plot(hc_ward,
     labels = rownames(house2),
     cex = 0.42,
     xlab = "",
     ylab = "Distance",
     main = "Euclidean Distance + Ward's Method")
rect.hclust(hc_ward, k = 6)
```

```{r clustering2-evaluate, echo=TRUE}
# Evaluate clustering: Euclidean + Average linkage (for comparison)
source("https://raw.githubusercontent.com/jreuning/sds363_code/refs/heads/main/HClusEval3.R.txt")
hclus_eval(house2,
           dist_m = 'euclidean',
           clus_m = 'average',
           plot_op = TRUE,
           print_num = 15)
```

## Clustering 3 – Maximum Distance & Complete Linkage

```{r clustering3-complete, echo=TRUE}
# Compute maximum (Chebyshev) distance matrix
dist_max <- dist(house2, method = "maximum")

# Perform hierarchical clustering using complete linkage
hc_max_complete <- hclust(dist_max, method = "complete")

# Plot dendrogram with 6 clusters highlighted
plot(hc_max_complete,
     labels = rownames(house2),
     cex = 0.42,
     xlab = "",
     ylab = "Distance",
     main = "Maximum Distance + Complete Linkage")
rect.hclust(hc_max_complete, k = 6)
```

```{r clustering3-evaluate, echo=TRUE}
# Evaluate clustering: Maximum distance + Complete linkage
source("https://raw.githubusercontent.com/jreuning/sds363_code/refs/heads/main/HClusEval3.R.txt")
hclus_eval(house2,
           dist_m = 'maximum',
           clus_m = 'complete',
           plot_op = TRUE,
           print_num = 15)
```

## Clustering 4 – Maximum Distance & Ward’s Method

```{r clustering4-ward, echo=TRUE}
# Compute maximum distance matrix
dist_max2 <- dist(house2, method = "maximum")

# Perform hierarchical clustering using Ward's method
hc_max_ward <- hclust(dist_max2, method = "ward.D")

# Plot dendrogram without evaluation for final grouping
plot(hc_max_ward,
     labels = rownames(house2),
     cex = 0.42,
     xlab = "",
     ylab = "Distance",
     main = "Maximum Distance + Ward's Method")

# Assign cluster memberships to data frame (6 clusters)
house$cluster <- cutree(hc_max_ward, k = 6)
```


## MANOVA and GLM

```{r manova-setup, echo=TRUE}
# Create binary indicators for deprivation, growth, and commuting
house$imd_rank     <- data$imd_rank
house$high_imd     <- ifelse(house$imd_rank <= median(house$imd_rank, na.rm = TRUE), 1, 0)
house$high_growth  <- ifelse(house$housepricechange >= median(house$housepricechange, na.rm = TRUE), 1, 0)
house$high_commuting <- ifelse(house$net_commuting >= median(house$net_commuting, na.rm = TRUE), 1, 0)

# Map region factor levels to descriptive names
house$region       <- factor(data$region[match(house$crime_rank, data$crime_rank)],
                             labels = c("East Midlands", "East of England", "London", 
                                        "North East", "North West", "South East", 
                                        "South West", "Wales", "West Midlands", 
                                        "Yorkshire and The Humber"))

# Inspect data structure and summary statistics
str(house)
summary(house)
```

```{r interaction-plots, echo=TRUE, warning=FALSE}
# Interaction plots for median house price and homeownership
interaction.plot(house$high_growth,
                 house$high_imd,
                 house$all_price_2015,
                 lwd = 3,
                 trace.label = "High IMD",
                 xlab = "High Growth",
                 main = "Interaction: Median House Price")

interaction.plot(house$high_growth,
                 house$high_imd,
                 house$households_owned,
                 lwd = 3,
                 trace.label = "High IMD",
                 xlab = "High Growth",
                 main = "Interaction: Home Ownership")
```

```{r glm-manova, echo=TRUE}
# Set contrasts for MANOVA
options(contrasts = c("contr.sum", "contr.poly"))

# Fit multivariate linear model for price and ownership
housemod <- lm(cbind(all_price_2015, households_owned) ~
                 high_imd * high_growth +
                 high_commuting +
                 region +
                 prop_health_vuln +
                 noqual +
                 prop_manu,
               data = house)

# Multivariate and univariate ANOVA (Type III sums of squares)
summary(Anova(housemod, type = 3), univariate = TRUE)

# Check residual multivariate normality
cqplot(housemod$residuals,
       label = "Residuals from GLM",
       main  = "Chi-Square Quantile Plot for GLM Residuals")
```

```{r glm-coefficients, echo=TRUE}
# Display coefficients for multivariate model
coef(housemod)
```

```{r contrasts, echo=TRUE}
# Reset to treatment contrasts for group comparison
options(contrasts = c("contr.treatment", "contr.poly"))

# Create London vs. Other grouping
house$region_group <- ifelse(house$region == "London", "London", "Other")

# Fit grouped model and test difference
housemod_grouped <- lm(cbind(all_price_2015, households_owned) ~
                         high_imd * high_growth +
                         high_commuting +
                         region_group +
                         prop_health_vuln +
                         noqual +
                         prop_manu,
                       data = house)

# Multivariate contrast: London vs. Other
linearHypothesis(housemod_grouped, "region_groupOther = 0")
